---
title: "Spatial causal inference in the presence of preferential sampling - Analysis on a pseudo-data set for reproducibility"
output:
  html_document:
    toc: true
    theme: united
---

## Load packages

Load packages to run the model.

```{r package}
library(spatstat)
library(fields)
library(geoR)
library(spam)
```


## Load relevant data and functions

Load the coordinates of grid cells and relevant functions for the analysis. To check reproducibility, set the seed number to be $1234$.

```{r load}
load("grid_pts08.RData") # grid cell coordinates
source("final_func.R")
set.seed(1234)
```

## Generate pseudo-data

Define the true values of the parameters.

```{r values}
alpha0 <- 7.0
alpha1 <- 8.0
beta <- c(-0.1, 9.0, 0.1, 1.0, -0.1, 0.6)

eta0 <- -3.0
eta1 <- -4.0
delta <- c(-1.0, -1.0, -0.5, -1.0, 2.0, -0.1)

phi0 <- 0.75
phi1 <- -0.75

gamma.u <- 0.5
gamma.v <- 1.0
```

Generate the spatial processes $\mathcal{W}$ by Gaussian processes with Matérn covariance kernels.

```{r process}
G <- nrow(dom.pts)
distMat <- rdist.earth(dom.pts, dom.pts, miles = FALSE)
diag(distMat) <- 0

U0.tilde <- t(chol(matern(distMat, 800, 0.5)))%*%rnorm(G, sd = sqrt(2.0))
U1.tilde <- t(chol(matern(distMat, 800, 0.5)))%*%rnorm(G, sd = sqrt(1.0))
V0.tilde <- t(chol(matern(distMat, 700, 0.5)))%*%rnorm(G, sd = sqrt(0.5))
V1.tilde <- t(chol(matern(distMat, 700, 0.5)))%*%rnorm(G, sd = sqrt(3.0))

U0 <- as.vector(U0.tilde)
U1 <- as.vector(U1.tilde + gamma.u*U0.tilde)
V0 <- as.vector(V0.tilde)
V1 <- as.vector(V1.tilde + gamma.v*V0.tilde)
```

Generate spatial covariates and define intensities for point processes given the covariates and the spatial processes.

```{r data}
Z1 <- t(chol(matern(distMat, 300, 0.5)))%*%rnorm(G)
Z2 <- t(chol(matern(distMat, 300, 0.5)))%*%rnorm(G)
Z3 <- t(chol(matern(distMat, 300, 0.5)))%*%rnorm(G)
Z4 <- t(chol(matern(distMat, 300, 0.5)))%*%rnorm(G)
Z5 <- t(chol(matern(distMat, 300, 0.5)))%*%rnorm(G)
Z6 <- t(chol(matern(distMat, 300, 0.5)))%*%rnorm(G)
Z <- cbind(Z1, Z2, Z3, Z4, Z5, Z6)
colnames(Z) <- paste("Variable", 1:6)

lambda0 <- exp(Z %*% delta + V0 + phi0*U0 + rnorm(G, 0, 0.1))
lambda1 <- exp(Z %*% delta + V1 + phi1*U1 + rnorm(G, 0, 0.1))
```

Generate Poisson random variables $N_{0,g}$ and $N_{1,g}$ per each grid cell for $g=1,\cdots,G$. 

```{r}
N0 <- rpois(n = G, lambda = exp(eta0)*lambda0) # sum(N0) = 1280
N1 <- rpois(n = G, lambda = exp(eta1)*lambda1) # sum(N1) = 1920
N <- N0 + N1
```

The realizations of the Poisson process in each grid cell represent the number of the survey sites in those cells. For simplicity, let us assume that the covariates of each survey site have the same values as those of the grid cell the site is located in. Define the treatment indicator $A(\mathbf{s})$ per each grid cell. Let the matrix $H$ denote the membership of the realized observations to the corresponding grid cells.

```{r}
X <- Z[rep(1:G, N),]

Tr <- c()
grid.ind <- c()
for(g in 1:G){
  Trg <- c(rep(0, N0[g]), rep(1, N1[g]))
  Tr <- c(Tr, Trg)
  grid.ind <- c(grid.ind, rep(g, N[g]))
}

H <- matrix(0, nrow = sum(N), ncol = G)
for(i in 1:sum(N)){
  H[i, grid.ind[i]] <- 1
}
H <- as(H, "sparseMatrix")
```

Let us generate the potential outcomes $Y_0(\mathbf{s})$ and $Y_1(\mathbf{s})$ using the site-level covariates and the spatial processes. Following the SUTVA assumption, generate the observed outcome as $Y(\mathbf{s})=Y_1(\mathbf{s})A(\mathbf{s})+Y(\mathbf{s})\left\{1-A(\mathbf{s})\right\}$.

```{r outcome}
Y0 <- alpha0 + X %*% beta + H%*%U0 + rnorm(sum(N), 0, sqrt(0.1))
Y1 <- alpha1 + X %*% beta + H%*%U1 + rnorm(sum(N), 0, sqrt(0.1))
Y <- as.vector(Y0*(1-Tr) + Y1*Tr)
triangle <- alpha1 - alpha0 + mean(U1 - U0)
```

Under the given data generation process, the true value of the average causal effect $\triangle$ is `r triangle`.

## Run the proposed model

Run the proposed model assuming $\boldsymbol{\beta}_0=\boldsymbol{\beta}_1=\boldsymbol{\beta}$ and $\boldsymbol{\delta}_0=\boldsymbol{\delta}_1=\boldsymbol{\delta}$.

```{r}
burn <- 10000
iters <- 50000
model <- Mat.PO.share(Y = Y, 
                      Tr = Tr, 
                      N0 = N0, 
                      N1 = N1, 
                      domain = dom.pts, 
                      X = X, 
                      Z = Z, 
                      H = H, 
                      center.X = FALSE, 
                      center.Z = FALSE, 
                      iters = iters, 
                      burn = burn, 
                      rho_mn = log(300), 
                      rho_sd = 1,
                      verbose = TRUE, 
                      update = 5000)
```

Now check whether the model converged or not. First, a visual inspection on the trace plot for the marginal posterior of $\triangle$ is done.

```{r}
plot(model$ATE[burn:iters], type = 'l') # Average causal effect
abline(h = triangle, col = 'red')
```

The trace plots for the marginal posterior distributions of $\phi_0$ and $\phi_1$ are as follows.

```{r}
plot(model$samps[burn:iters, "phi0"], type = 'l') # phi0
abline(h = phi0, col = 'red')
```

```{r}
plot(model$samps[burn:iters, "phi1"], type = 'l') # phi1
abline(h = phi1, col = 'red')
```


The mean posterior values of the coefficients corresponding to the covariates in the potential outcome models are as follows.

```{r}
colMeans(model$samps[burn:iters, paste("beta", colnames(X))])
beta 
```

The mean posterior values of the coefficients corresponding to the covariates in the point process models are as follows.

```{r}
colMeans(model$samps[burn:iters, paste("delta", colnames(X))])
delta
```

